---
title: "Cyclist - Analysis & Review"
author: "Ryan Kroker"
date: "2023-06-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Environment Setup 

###### ________________________  

In this data analysis we will be using the following packages and libraries:  
```{r libraries, echo=TRUE}
library(tidyverse)
library(lubridate)
library(ggplot2)
```  
  
The following CSV files have been read into data frames:  

```{r load, echo=TRUE}
q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")
```  
  
### Prepare and Explore  

###### __________________________  
  
Let's review the initial data with the business questions and objectives in mind.

###### Column Names:  

```{r explore, echo=TRUE}
colnames(q3_2019)
colnames(q4_2019)
colnames(q2_2019)
colnames(q1_2020)
```

Upon initial review, we can see that the Columns do not match for each data sets, will need to clean to make them consistent so we can merge them into a single data frame or "stacked" data frame for a comprehensive comparison.  

###### Rename Columns:    
```{r rename_col, echo=TRUE}
q4_2019 <- rename(q4_2019, ride_id = trip_id, rideable_type = bikeid, started_at = start_time, ended_at = end_time, start_station_name = from_station_name, start_station_id = from_station_id, end_station_name = to_station_name, end_station_id = to_station_id, member_casual = usertype)

q3_2019 <- rename(q3_2019, ride_id = trip_id, rideable_type = bikeid, started_at = start_time, ended_at = end_time, start_station_name = from_station_name, start_station_id = from_station_id, end_station_name = to_station_name, end_station_id = to_station_id, member_casual = usertype)

q2_2019 <- rename(q2_2019, ride_id = "01 - Rental Details Rental ID", rideable_type = "01 - Rental Details Bike ID", started_at = "01 - Rental Details Local Start Time", ended_at = "01 - Rental Details Local End Time", start_station_name = "03 - Rental Start Station Name", start_station_id = "03 - Rental Start Station ID", end_station_name = "02 - Rental End Station Name", end_station_id = "02 - Rental End Station ID", member_casual = "User Type")

```
  
Let's reinspect the data to ensure the columns now match, as well as any data types that might be misaligned.  
  
```{r inspect_dfs, echo=TRUE}
str(q1_2020)
str(q4_2019)
str(q3_2019)
str(q2_2019)
```
  
Upon further inspection, we can see Ride_id and Rideable_type are different data types, will need to convert to same type. We will choose the CHR data type, as the latest 2020 data uses this data type, and it is easier to manipulate for our purposes.  
  
###### Convert data types  
```{r convert, echo=TRUE}
q4_2019 <-  mutate(q4_2019, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type)) 
q3_2019 <-  mutate(q3_2019, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type)) 
q2_2019 <-  mutate(q2_2019, ride_id = as.character(ride_id), rideable_type = as.character(rideable_type)) 
```
  
###### Stack Quarter frames into one Year frame  
  
```{r merge, echo=TRUE}
all_trips <- bind_rows(q2_2019, q3_2019, q4_2019, q1_2020)
```
  
###### Inspect new frame all_trips  
  
```{r inspect_AT, echo=TRUE}
glimpse(all_trips)
```
  
There are a few fields not used or with no data, gender, birthyear, start_lat, start_lng, end_lat, end_lng, 01 - Rental Details Duration in Seconds, 05 - Member Details Birth Year. These can be removed.  
  
###### Remove unused columns  
  
```{r remove_col, echo=TRUE}
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "01 - Rental Details Duration In Seconds Uncapped", "05 - Member Details Member Birthday Year", "Member Gender", "tripduration"))

```
  
### Correct and Add Data  

###### ______________________________  
  
Inspect all_trips  
  
List of column names  

```{r colnames_AT, echo=FALSE}
colnames(all_trips)
```
  
How many rows are in the data frame?  
  
```{r num_rows_AT, echo=FALSE}
nrow(all_trips)
```
  
Dimensions of the data frame?  
  
```{r dimensions_AT, echo=FALSE}
dim(all_trips)
```
  
Preview Data  
  
```{r glimpse_AT, echo=FALSE}
glimpse(all_trips)
```
  
Statistical Summary  
  
```{r sum_AT, echo=FALSE}
summary(all_trips)
```
  
##### Identified Problems:  
  
There are a few problems we will need to fix:  

1. In the "member_casual" column, there are two names for members ("member" and "Subscriber") and two names for casual riders ("Customer" and "casual"). We will need to consolidate that from four to two labels.  
2. The data can only be aggregated at the ride-level, which is too granular. We will want to add some additional columns of data -- such as day, month, year -- that provide additional opportunities to aggregate the data.  
3. We will want to add a calculated field for length of ride since the 2020Q1 data did not have the "tripduration" column. We will add "ride_length" to the entire dataframe for consistency.  
4. There are some rides where tripduration shows up as negative, including several hundred rides where Divvy took bikes out of circulation for Quality Control reasons. We will want to delete these rides.  
  
##### 1. Let's begin by seeing how many observations fall under each usertype  

```{r member_check, echo=TRUE}
table(all_trips$member_casual)
```
  
Reassign Subscriber and Customer to the desired values (we will go with the latest 2020 labels)  
  
```{r recode_member_types, echo=TRUE}
all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual, "Subscriber" = "member", "Customer" = "casual"))
```
  
Check to make sure the proper number of observations were reassigned  
  
```{r member_check2, echo=TRUE}
table(all_trips$member_casual)
```
  
  
##### 2. Add columns for date, month, day and year of each ride (allows for different levels of aggregation)  
  
```{r add_col, echo=TRUE}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```
  
Let's add a "ride_length" calculation to all_trips (in seconds) so we can easily compare these metrics later. 
  
```{r ride_length, echo=TRUE}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```
  
  
##### 3. The new column ride_length needs to be converted to numeric so calculations can be performed  

Conversion to Number  
  
```{r convert_num, echo=TRUE}
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
```
  
  
##### 4. The new data frame inlcudes data that is either incomplete or unessacary for our analisys. Let's remove that "bad" data so we can focus on the important data.  
  
Create new data frame without "bad" data  

```{r remove_bad_data, echo=TRUE}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```
  
  
### Conduct Descriptive Analysis  
  
###### ______________________________  

#### Descriptive analysis on ride_length (all figures in seconds)  
  
Straight Average (total ride length / rides)  

```{r rl_mean, echo=FALSE}
mean(all_trips_v2$ride_length)
```
  
Midpoint number in the ascending array of ride lengths  
  
```{r rl_median, echo=FALSE}
median(all_trips_v2$ride_length)
```
  
Longest Ride  
  
```{r rl_max, echo=FALSE}
max(all_trips_v2$ride_length)
```
  
Shortest Ride  
  
```{r rl_min, echo=FALSE}
min(all_trips_v2$ride_length)
```
  
Ride Length Summary  
  
```{r rl_summary, echo=FALSE}
summary(all_trips_v2$ride_length)
```
  
#### Compare members and casual users (mean, median, maximum, and minimum ride lenths) 
  

```{r agg_membertype, echo=TRUE}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```
  
#### Average time each day for members/casuals  
  
```{r agg_time_per_day, echo=FALSE}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```
  
The days are out of order, and makes this difficult to compare weekends and weekdays. Let's correct the order of the week for a more comprehensive analysis.  
  
###### Correcting days of the week order  
```{r sort_days, echo=TRUE}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```
  
###### Recheck time each day for members/casuals  
  
```{r agg_time_per_day_check, echo=FALSE}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```
  
####  Ridership data by type and weekday comparison  
  
```{r ridership_summary, echo=FALSE}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(member_casual, weekday)	
```
  
#### Visualize number of rides by rider type  
  
```{r ridership_viz, echo=FALSE}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```
  
Members have a higher number of rides per day then our Casual riders, especially during weekdays.  
  
  * What could we do to increase the number of rides during the weekdays for our casual riders?  
  
  
#### Visualize Average Duration  
  
```{r avg_duration_viz, echo=FALSE}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(),average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +geom_col(position = "dodge")

```
  
Casual riders on average take longer rides then Members, regardless of the day of the week. 
  
  
#### Review  
  
The data provides tells us that while on average, Members have a higher amount of rides per day, though each ride is shorter then the average casual member. This could mean that Members enjoy shorter more frequent rides, while casual customers would rather take a single longer ride each day.  

  * To attract more Members, I would recommend Cyclist may want to implement incentives for longer rides as it may encourage casual riders to become full members, it may also incentives existing members to take longer rides.  

  * To maintain Members, I would recommend Cyclist celebrate the number of rides of their existing members. This may also encourage casual riders to increase the frequency of their rides.  
  
  * In general, I would recommend that Cyclist implements an incentive system that recognizes its members for their achievements in cycling. Number of rides, ride length, and history with Cyclist would all be great metrics to start with.  
  

    
##### Export Files  

We've spent some time making new calculations, lets reserve that data for further analysis. We will export this data in a CSV for further review named "ave_ride_length.csv".  
